Emittance preservation of an electron beam in a loaded quasi-linear plasma wakefield

### 2017-09-01 Tex/Paper.tex
Edits after meeting with Erik for section III

%~ The first run of the experiment is set up with a $10\unit{m}$ rubidium vapour cell.
%~ The rubidium vapour is
%~ ionised by a $780\unit{nm}$, $4.5\unit{TW}$ peak power laser with a pulse length of $100−120\unit{fs}$.
The AWAKE experiment, currently in its first stages of operation at CERN, uses the SPS beam at $400\unit{GeV}$ as its
driver, and a nominal plasma density $7\nexp{14}\unit{cm}^{-3}$ \cite{gschwendtner:2016}. This plasma density
corresponds to a plasma wavelength $\lambda_{pe} = 1.26\unit{mm}$. The plasma density is matched to the SPS proton beam
such that $k_{pe}\sigma_{r} = 1$ where $k_{pe} = 2\pi/\lambda_{pe}$ is the plasma wave number. The aim of the first
phase of Run 1 of the experiment is to demonstrate self-modulation of the proton beam, and in 2018 to sample the
wakefield with a long electron beam. The study presented here is for Run 2 \cite{adli:2016}, which aims to demonstrate
acceleration of a short electron beam to high energy with a minimal increase in emittance and energy spread.

@@

%~ The beam matching relation is given by
%~ \begin{align}
    %~ \beta = \frac{\sigma_r^2}{\epsilon_g} = \frac{\lambda_pe}{2\pi}\sqrt{2\gamma_{rel}}, \label{EQ:Matched}
%~ \end{align}
%~ where $\lambda_{p}$ is the plasma wavelength. For a reasonable initial emittance of $2\unit{\mu m}$, this corresponds to
%~ a $\sigma_{r}$ of $5.25\unit{\mu m}$ -- a very narrow bunch compared to the drive beam $\sigma_{r} = 200\unit{\mu m}$.
%~ The charge density of this compact witness bunch reaches that of the plasma density at only a few $\unit{pC}$, but
%~ reaches optimal beam loading at around $1-200\unit{pC}$. The implication here being that the witness bunch produces its
%~ own non-linear wakefield already at the head of the bunch. The majority of the electrons within it will therefore see a
%~ linear focusing force preventing the bulk of the beam from undergoing emittance growth.

@@

The peak density of the simulation drive beam is $0.83\cdot n_{0}$, producing a quasi-linear wakefield. The single bunch
setup uses the baseline proton energy $W_{pb} = 400\unit{GeV}$ and transverse size $\sigma_{r} = 200\unit{\mu m}$. We
set the length of the drive beam to $\sigma_{z} = 40\unit{\mu m}$. In order to create a stable environment to study the
evolution of the witness beam we prevented the proton beam from evolving radially by increasing the particle mass by a
factor of $1\mexp{6}$. We are not considering the evolution of the proton beam itself in this study.

@@

In these 3D simulations we have used a matched beam for an initial emittance of $2\unit{\mu m}$, corresponding to a
$\sigma_{r}$ of $5.25\unit{\mu m}$. The beam matching relation is given by
\begin{align}
    \beta = \frac{\sigma_r^2}{\epsilon_g} = \frac{\lambda_pe}{2\pi}\sqrt{2\gamma_{rel}}, \label{EQ:Matched}
\end{align}
where $\lambda_{p}$ is the plasma wavelength. This is a very narrow beam compared to the drive beam
$\sigma_{r} = 200\unit{\mu m}$. The charge density of this compact witness beam reaches that of the plasma density at
only a few $\unit{pC}$, but reaches optimal beam loading at around $1-200\unit{pC}$. The implication here being that the
witness beam produces its own non-linear wakefield already at the head of the beam. The majority of the electrons
within it will therefore see a linear focusing force preventing the bulk of the beam from undergoing emittance growth.

@@

@@

The single drive beam setup is designed to behave similarly to the self-modulated case. However, since the drive beam is
prevented from significant transverse evolution, we are presented with an idealised case where the electron witness beam
sees consistent wakefields throughout the plasma stage. The $E_{z}$-field generated by the proton drive beam is seen as
the blue line in figure \ref{Fig:BeamLoading}, shown with and without the electron beam present. With a proton beam
density $n_{pb} \simeq n_{0}$, we are in the quasi-linear regime but near non-linear \cite{rosenzweig:2010}. There is
some depletion of plasma electrons, but not enough to form a bubble. The dashed green line in the lower part of figure
\ref{Fig:BeamLoading} shows that the on-axis plasma density has a depletion of $67\%$, close to what we see in full
scale reference simulations for AWAKE Run 2 \cite{awake_collaboration:2016}.

@@

The witness beam generates its own wakefield, which in the accelerating phase partially cancels out the $E_{z}$ field
generated by the drive beam. For a finite length beam this causes the particles in the tail of the witness beam to be
accelerated less than those at the front resulting in higher energy spread in exchange for higher energy transfer
\cite{van_der_meer:1985}. Since $E_{z} \propto \rho_{b}$, a beam profile $\rho_{b}(\xi)$ that perfectly flattens the
accelerating field within the witness beam will reduces energy spread to zero. The ideal shape is triangular or
trapezoidal with the peak density in the direction of the beam velocity \cite{katsouleas:1987, tzoufras:2009}.

@@

For Gaussian beams a reasonable flat field, and consequently low energy spread, can be achieved by controlling the
charge density to achieve a similar effect. This, however, puts the front of the witness beam outside of the ideal
region while the middle bulk of the beam has a trapezoidal-like shape. In our case, with a matched beam to the plasma
density defined by equation \ref{EQ:Matched}, we are limited on how much total charge we can accelerate without reaching
charge densities that will overload the field. The test case shown in figures \ref{Fig:BeamLoading} and
\ref{Fig:PlasmaDenTWake} has a total charge of $100\unit{pC}$ with a $\sigma_{r}=5.23\unit{\mu m}$ matching an initial
normalised emittance $\epsilon_{0} = 2\unit{\mu m}$.

As illustrated in figure \ref{Fig:BeamPS} our base case of a $100\unit{pC}$, $60\unit{\mu m}$ long witness beam, as
expected, shows a difference in energy transfer to the tail of the beam compared to the centre where the $E_{z}$-field
is nearly flat. Also the front of the beam has a large tail in momentum space as this regions sees a consistently
smaller accelerating field. The positioning of the witness beam is chosen to put the bulk of the charge as close to the
peak accelerating field as possible.

Due to the high charge density resulting from a very narrow beam, the witness beam's own wakefield reaches the full
non-linear regime with $n_{eb} \approx 35\times n_{0}$. This second wakefield is driven by the beam's own front. The
space charge rapidly becomes high enough to start expel plasma electrons from the beam axis, and form the characteristic
electrons sheet that defines the blowout regime. As a result the plasma region reaches full depletion close to the peak
of the electron beam, which in turn generates strong focusing fields, with gradients of $20\unit{kT/m}$ near the beam
axis (see figure \ref{Fig:PlasmaDenTWake}).

@@

The witness beam's self-generated bubble prevents the tailing end of the beam to see any significant emittance growth.
For the majority of the cases we studied that maintained a stable accelerating structure, about $70-80\%$ of the beam
retained its initial emittance. Figure \ref{Fig:BeamEmitt} shows the $100\unit{pC}$ and $\sigma_{z} = 60\unit{\mu m}$
case extended to a $100\unit{m}$ plasma and the drive beam energy increased to $7\unit{TeV}$ (LHC energy) to prevent
de-phasing. De-phasing for the SPS beam case starts to become significant after about $50\unit{m}$.

The on-axis density of the electron beam increases as its gamma factor increases and its transverse size decreases. The
beam radius follows the evolution given by equation \ref{EQ:Matched} with $\lambda_{pe} = \lambda_{0}$ for the section
of the beam where normalised emittance is preserved. This has the potential to cause overloading of the field. However,
our $100\unit{pC}$ case is slightly under loaded, and this does not seem to be a significant effect in this instance.

@@

Our configuration also shows some robustness to small electron beam transverse offsets as the bubble generated by the
witness beam wakefield follows the beam itself. The head of the beam does not benefit from this effect, and we see
some defocusing in this region. This effect is likely to be greater for larger offsets as the beam oscillates around
the axis of the drive beam wakefield (see right hand plots of figure \ref{Fig:BeamFilament}). However, since the proton
wakefield creates a quasi linear condition, we still see the head of the beam stabilising after some time. The off-axis
case has a larger initial emittance growth (see left hand plots of figure \ref{Fig:BeamFilament}), but the growth levels
off after the first few metres and stays constant for the remainder of the plasma accelerator section (figure
\ref{Fig:BeamEmitt}).

>>>>

%~ The first run of the experiment is set up with a $10\unit{m}$ rubidium vapour cell.
%~ The rubidium vapour is
%~ ionised by a $780\unit{nm}$, $4.5\unit{TW}$ peak power laser with a pulse length of $100−120\unit{fs}$.
The AWAKE experiment, currently in its first stages of operation at CERN, uses the SPS beam at $400\unit{GeV}$ as its
driver, and a nominal plasma density $7\nexp{14}\unit{cm}^{-3}$ \cite{gschwendtner:2016}. This plasma density
corresponds to a plasma wavelength $\lambda_{pe} = 1.26\unit{mm}$. The plasma density is matched to the SPS proton beam
such that $k_{pe}\sigma_{x} = 1$ where $k_{pe} = 2\pi/\lambda_{pe}$ is the plasma wave number. The aim of the first
phase of Run 1 of the experiment is to demonstrate self-modulation of the proton beam, and in 2018 to sample the
wakefield with a long electron beam. The study presented here is for Run 2 \cite{adli:2016}, which aims to demonstrate
acceleration of a short electron beam to high energy with a minimal increase in emittance and energy spread.

@@

%~ The beam matching relation is given by
%~ \begin{align}
    %~ \beta = \frac{\sigma_r^2}{\epsilon_g} = \frac{\lambda_pe}{2\pi}\sqrt{2\gamma_{rel}}, \label{EQ:Matched}
%~ \end{align}
%~ where $\lambda_{p}$ is the plasma wavelength. For a reasonable initial emittance of $2\unit{\mu m}$, this corresponds to
%~ a $\sigma_{x}$ of $5.25\unit{\mu m}$ -- a very narrow bunch compared to the drive beam $\sigma_{x} = 200\unit{\mu m}$.
%~ The charge density of this compact witness bunch reaches that of the plasma density at only a few $\unit{pC}$, but
%~ reaches optimal beam loading at around $1-200\unit{pC}$. The implication here being that the witness bunch produces its
%~ own non-linear wakefield already at the head of the bunch. The majority of the electrons within it will therefore see a
%~ linear focusing force preventing the bulk of the beam from undergoing emittance growth.

@@

The peak density of the simulation drive beam is $0.83\cdot n_{0}$, producing a quasi-linear wakefield. The single bunch
setup uses the baseline proton energy $W_{pb} = 400\unit{GeV}$ and transverse size $\sigma_{x} = 200\unit{\mu m}$. We
set the length of the drive beam to $\sigma_{z} = 40\unit{\mu m}$. In order to create a stable environment to study the
evolution of the witness beam we prevented the proton beam from evolving radially by increasing the particle mass by a
factor of $1\mexp{6}$. We are not considering the evolution of the proton beam itself in this study.

@@

In these 3D simulations we have used a matched beam for an initial emittance of $2\unit{\mu m}$, corresponding to a
$\sigma_{x}$ of $5.25\unit{\mu m}$. The beam matching relation is given by
\begin{align}
    \beta = \frac{\sigma_r^2}{\epsilon_g} = \frac{\lambda_pe}{2\pi}\sqrt{2\gamma_{rel}}, \label{EQ:Matched}
\end{align}
where $\lambda_{p}$ is the plasma wavelength. This is a very narrow beam compared to the drive beam
$\sigma_{x} = 200\unit{\mu m}$. The charge density of this compact witness beam reaches that of the plasma density at
only a few $\unit{pC}$, but reaches optimal beam loading at around $1-200\unit{pC}$. The implication here being that the
witness beam produces its own non-linear wakefield already at the head of the beam. The majority of the electrons
within it will therefore see a linear focusing force preventing the bulk of the beam from undergoing emittance growth.

@@

In the following section we study a witness beam with a total charge of $100\unit{pC}$, a $\sigma_{z}=60\unit{\mu m}$,
and with a $\sigma_{x}=5.23\unit{\mu m}$ matching an initial normalised emittance $\epsilon_{0} = 2\unit{\mu m}$. We
refer to this parameter set as our base case.

@@

The single drive beam setup is designed to behave similarly to the self-modulated case. However, since the drive beam is
prevented from significant transverse evolution, we are presented with an idealised case where the electron witness beam
sees consistent wakefields throughout the plasma stage. The $E_{z}$-field generated by the proton drive beam is seen as
the blue line in figure \ref{Fig:BeamLoading}, shown with and without the electron beam present. With a proton beam
density $n_{pb} \simeq n_{0}$, we are in the quasi-linear regime but near non-linear \cite{rosenzweig:2010}. The dashed
green line in the lower part of figure \ref{Fig:BeamLoading} shows that the on-axis plasma density has a depletion of
$67\%$, close to what we see in full scale reference simulations for AWAKE Run 2 \cite{awake_collaboration:2016}.

@@

%~ The witness beam generates its own wakefield, which in the accelerating phase partially cancels out the $E_{z}$ field
%~ generated by the drive beam. For a finite length beam this causes the particles in the tail of the witness beam to be
%~ accelerated less than those at the front resulting in higher energy spread in exchange for higher energy transfer
%~ \cite{van_der_meer:1985}. A beam profile $\rho_{b}(\xi)$ that perfectly flattens the
%~ accelerating field within the witness beam will reduces energy spread to zero. The ideal shape is triangular or
%~ trapezoidal with the peak density in the direction of the beam velocity \cite{katsouleas:1987, tzoufras:2009}.

The witness beam generates its own wakefield which partially cancels out the $E_{z}$ field generated by the drive beam.
With an ideally shaped electron beam charge profile it is possible to optimally load the field, such that the
accelerating $E_{z}$ is constant along the bunch \cite{katsouleas:1987, tzoufras:2009}.

@@

While the ideal shape is trapezodial, for Gaussians beams a reasonable flat field, and consequently low energy spread,
can be achieved by controlling the charge density to achieve a similar effect. This, however, puts the front of the
witness beam outside of the ideal region while the middle bulk of the beam has a trapezoidal-like shape. In our case,
with a matched beam to the plasma density defined by equation \ref{EQ:Matched}, we are limited on how much total charge
we can accelerate without reaching charge densities that will overload the field.

As illustrated in figure \ref{Fig:BeamPS} our base case shows a difference in energy transfer to the tail of the beam
compared to the centre where the $E_{z}$-field is nearly flat. Also the front of the beam has a large tail in momentum
space as this regions sees a smaller accelerating field. The positioning of the witness beam is chosen to put the bulk
of the charge as close to the peak accelerating field as possible.

The charge density required to optimally load the $E_{z}$ field is $n_{eb} \approx 35\times n_{0}$. This means that the
witness beam's own wakefield is in the fully non-linear regime, where the space charge force is sufficient to blow-out
all plasma electrons resulting in an ion column. This ion column, as is well known \cite{rosenzweig:1991}, provides a
linear focusing force on the part the electron beam within the ion column, and therefore prevents emittance growth for
this region. This effect is shown for our base case in figure \ref{Fig:PlasmaDenTWake}. The focusing field has a
gradient of $20\unit{kT/m}$ near the beam axis.

%~ Due to the high charge density resulting from a very narrow beam, the witness beam's own wakefield reaches the full
%~ non-linear regime with $n_{eb} \approx 35\times n_{0}$. This second wakefield is driven by the beam's own front. The
%~ space charge rapidly becomes high enough to start expel plasma electrons from the beam axis and form the characteristic
%~ electrons sheet that defines the blowout regime. As a result the plasma region reaches full depletion close to the peak
%~ of the electron beam, which in turn generates strong focusing fields, with gradients of $20\unit{kT/m}$ near the beam
%~ axis (see figure \ref{Fig:PlasmaDenTWake}).

@@

%~ The witness beam's self-generated bubble prevents the tailing end of the beam to see any significant emittance growth.
For the majority of the cases we studied that maintained a stable accelerating structure, about $70-80\%$ of the beam
retained its initial emittance. Figure \ref{Fig:BeamEmitt} shows emittance growth for the base case as a function of
propagation length. No significant emittance growth was observed for propagation lengths up to $100\unit{m}$. The drive
beam energy increased to $7\unit{TeV}$ (LHC energy) to prevent de-phasing. De-phasing for the SPS beam case starts to
become significant after about $50\unit{m}$.

The on-axis density of the electron beam increases as its gamma factor increases and its transverse size decreases. The
beam radius follows the evolution given by equation \ref{EQ:Matched} with $\lambda_{pe} = \lambda_{0}$ for the section
of the beam where normalised emittance is preserved as shown in figure \ref{Fig:BeamFilament}. This has the potential to
cause overloading of the field. However, for our base case this effect is not significant.

@@

%~ Our configuration also shows some robustness to small electron beam transverse offsets as the bubble generated by the
%~ witness beam wakefield follows the beam itself.

Since the accelerated electron beam creates its own focusing bubble, the emittance of the part of the beam inside the
bubble is not affected by small beam offsets with respect to the proton beam axis. This is an added benefit of this new
accelerating regime, and may ease the transverse injection tolerances. The head of the beam does not benefit from this
effect, but since the proton beam creates a quasi linear wake, the head of the beam still stabilises after some time.
This is illustrated on figure \ref{Fig:BeamEmitt} for an electron beam offset of $1\sigma_{x}$. In this case there is a
larger initial emittance growth (see left hand plots of figure \ref{Fig:BeamFilament}), but the emittance growth stop
after the first few metres (figure \ref{Fig:BeamEmitt}). This effect is likely to be greater for larger offsets as the
beam oscillates around the axis of the drive beam wakefield (see right hand plots of figure \ref{Fig:BeamFilament}).
